---
title: "Preparing the spatial data"
output: html_notebook
---

The accessibility analysis also depends on the spatial data. In this post, I will load the `shapefile` and process the euclidean distance among the centroids (point at the geometric center of the geometry). There are several packages and ways to do this process. For now, I will focus on the `sf` package. Later we will test other possibilities.

```{r loading library and shp}
library(sf)
shp <- read_sf('../data/Zonas_2017_region.shp')
head(shp)
```
## Getting the centroids

First, let's take a quick look (without embelishments) at the `shapefile`.

```{r}
plot(shp[,'geometry'])
```
In the `sf` (simple features) package, we can apply the function `st_centroid` to get the centroids of the geometry.

```{r centroids}
centroids <- st_centroid(shp)
plot(centroids[,'geometry'])
```
See that instead of areas/regions, we have points now.

## Processing the distances

Having the centroids, a simple way to include the spatial data to analyze the urban accessibility is through the **euclidean distance**. We can create the matrix by using the `st_distance` of the `sf` library.

```{r processing distance-matrix}
distance <- st_distance(centroids, centroids)

write.csv(distance, '../data/distance_matrix.csv')
```


